% !TEX root = ../main.tex
\chapter[Generalized Algebraic Theories]{Generalized Algebraic Theories}

The goal of this chapter is give meaning to the concept of 
term, type and theory, while introducing some tools that we will
use later when we present the algorithm.

By theories we mean Generalized Algebraic Theories (or GATs), 
introduced by John Cartmell in his thesis \cite{cartmell1978} \cite{cartmell1986}.
GATs give us both a very idiomatic way of presenting
theories and powerful interpretation tools like
Context Categories, introduced later in this section.
The notion of generalized algebraic theory is a generalization of
the notion of many-sorted algebraic theory in the following manner:
while in Martin L{\"o}ff's many-sorted algebraic theories sorts
are constant types, interpreted as sets, in generalized
algebraic theories sorts might be constant or variable,
over a specified range of types, and variable types
are interpreted as families of sets. So a generalized
algebraic theory consists of:
\begin{enumerate}
    \item a set of sorts, that can be constant or variable;
    \item a set of operators, with the sort of the arguments and value specified;
    \item a set of axioms, where each axiom is an identity between similar expressions. 
\end{enumerate}
Let's present category theory for example:
The set of sorts contains two elements $Ob$ and $Hom$.
$Ob$ is a constant type and $Hom$ is a variable type,
for a pair of terms of type $Ob$, name them $x, y$, we
have a type $Hom(x, y)$.
Here we see a key point in general algebraic theories,
types vary over terms, no over types.
The set of operators also has two elements $id$ and $\circ$.
The $id$ operator has one argument $x$ of type $Ob$,
and returns a value of type $Hom(x, x)$. This shows
how the type of value of an operator might vary depending
on the arguments. 
The $\circ$ operator (or composition) expects 5 arguments, 
three of type $Ob$, name them $x, y, z$, one of type $Hom(x, y)$,
say $f$, and one of type $Hom(y, z)$, named $g$, and it returns
a value of type $Hom(x, z)$.
Here we see that the type of an argument might also vary depending
on a previous argument.
{\color{red}Informally, instead of writing $\circ(x, y, z, f, g)$ we can
write $\circ(f, g)$, as the first three arguments might be inferred from the last two.
This notation will be discussed in a further section.}

Now we will introduce the notation we will be using when talking 
about terms and types. {\color{orange} This notation will be formally defined
in a later section.} So consider you have a set of variables $V$,
we will associate terms to variables as required.
When asserting that a term $t$ is of type $\Delta$ we will write
$t : \Delta$, but if $t$ has variables $x_1, \dots, x_n$ occurring in it this assertion
does not make sense unless we assign some types to $x_1, \dots, x_m$.
To do this we will write:
\begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n: \Delta_n
    }{
        t : \Delta
    }
\end{equation*}
that reads as: if $x_1$ is a variable of type $\Delta_1$, \dots and if $x_n$ is a variable of type $\Delta_n$,
then $t$ is a term of type $\Delta$. 
Similarly, if we want to assert that if $x_1$ is a variable of type $\Delta_1$,
\dots and if $x_n$ is a variable of type $\Delta_n$,
then $\Delta$ is a type, we can write:
\begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n : \Delta_n 
    }{
        t : \Delta
    }
\end{equation*}
this assertions are called rules.
Rules are used to express which expressions are well-formed.
We will work with this rules as units instead of the base expressions.
This is because terms and types do not make sense by themselves
but they need to be thought of in a context. 

Axioms are also written as rules. The conclusion of those rules
are equalities between types or between terms of the same type.
We write those rules as:
\begin{minipage}{0.5\textwidth}
  \begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n : \Delta_n
    }{
        \Delta = \Delta'
    }
  \end{equation*}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
  \begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n : \Delta_n
    }{
        t = t' : \Delta
    }
  \end{equation*}
\end{minipage}
and read as: if $x_1$ is a variable of type $\Delta_1$, \dots and $x_n$ is a variable of type $\Delta_n$,
then $\Delta = \Delta'$, or $t = t'$ as terms of type $\Delta$, respectively.

% TODO: fix alignment
To continue with the previous example, we can write the axioms
of category theory as:
\begin{alignat*}{2}
    &x, y : Ob, f : Hom(x, y),  \quad  \circ(id(x), f) = f \\
    &x, y : Ob, f : Hom(x, y),  \quad  \circ(f, id(y)) = f \\
    &w, x, y, z : Ob, f : Hom(w, x), g : Hom(x, y), h : Hom(y, z), \\ &\qquad\circ(\circ(f, g), h) = \circ(f, \circ(g, h))
\end{alignat*}

When presenting a theory's language i.e. the sets of sorts and operators,
we must present each symbol with a rule specifying the role
of this symbol, either as a sort symbol or as specifically typed operator symbol. 
This rules are called introductory rules.
In the case of a sort symbol $A$ the rule should be like 
\begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n : \Delta_n
    }{
        A(x_1, \dots, x_n) \text{ is a type}
    }
\end{equation*}
clearly specifying the types $A$ is dependent on.
In the case of an operator symbol, a rule of the form
\begin{equation*}
    \dfrac{
        x_1 : \Delta_1, \dots, x_n : \Delta_n
    }{
        f(x_1, \dots, x_n) : \Delta
    }
\end{equation*}
suffices to specify the types of the arguments and the value of $f$.

Finally, to present a theory we will give a set os symbols,
each associated to its introductory rule, and a set of axioms.
And, of course, everything must be well-formed, {\color{orange} something we 
will address in the following section.}
So now we can give a formal presentation of the theory of categories:

\begin{tabular}{ll}
\textit{Symbol} & \textit{Introductory Rule} \\[6pt]
Ob  & Ob is a type. \\[4pt]
Hom & $x, y \in \mathrm{Ob} : \mathrm{Hom}(x, y)$ is a type. \\[4pt]
$\circ$ & $x, y, z \in \mathrm{Ob},\, f \in \mathrm{Hom}(x, y),\, g \in \mathrm{Hom}(y, z) : \circ(f, g) \in \mathrm{Hom}(x, z).$ \\[4pt]
id  & $x \in \mathrm{Ob} : \mathrm{id}(x) \in \mathrm{Hom}(x, x).$ \\
\end{tabular}

\begin{tabular}{ll}
\textit{Axioms} \\[6pt]
$\circ(\mathrm{id}(x), f) = f,$ & whenever $x, y \in \mathrm{Ob}$ and $f \in \mathrm{Hom}(x, y).$ \\[4pt]
$\circ(f, \mathrm{id}(y)) = f,$ & whenever $x, y \in \mathrm{Ob}$ and $f \in \mathrm{Hom}(x, y).$ \\[4pt]
$\circ(\circ(f, g), h) = \circ(f, \circ(g, h)),$ 
  & whenever $w, x, y, z \in \mathrm{Ob},\, f \in \mathrm{Hom}(w, x),$ \\
  & \hspace{4.5em} $g \in \mathrm{Hom}(x, y)$ and $h \in \mathrm{Hom}(y, z).$ \\
\end{tabular}

\section{Formal Definition}

In this section we will formally define all of the concepts 
we just introduced.
To give a notion of well-formed introductory rule, we need the concept of drivability,
which also depends on the introductory rules. 
To tackle this problem we will leave aside the question of
well-formedness until we have the complete set of derived rules
of a theory.
We will first define a pretheory and then define a theory
as a well-formed pretheory.

We will assume to have countable set of variables $V$.
Lets define the notion of rule on an alphabet $W$,
the elements of $W$ are called symbols.

\begin{definition}
    The set of \textbf{expressions} is defined inductively
    such that expressions are finite sequences of elements
    of $W \cup V \cup \{(\} \cup \{)\} \cup \{,\}$
    defined by the clauses:
    \begin{enumerate}
        \item if $x \in V$, then $x$ is an expression;
        \item if $L \in W$, then $L$ is an expression;
        \item if $L \in W$ and $e_1, \dots, e_n$ are expressions,
              then $L(e_1, \dots, e_n)$ is an expression.
    \end{enumerate}
\end{definition}

Now we can define the building blocks of a rule:
\begin{definition}
    A \textbf{premise} is defined to be a sequence of $V \times$ the set of expressions.
    The empty sequence is allowed and we call it an empty premise.
    If $((x_1, \Delta_1), \dots, (x_n, \Delta_n))$ is a premise,
    we write it as $x_1 : \Delta_1, \dots, x_n : \Delta_n$.
\end{definition}
\begin{definition}
    We call a \textbf{conclusion} to one of the following:
    \begin{enumerate}
        \item $T$-conclusion: defined by a single expression $\Delta$
              and written as '$\Delta$ is a type'.
        \item $\in$-conclusion: defined by a pair of expressions $(t, \Delta)$
              and written as '$t \in \Delta$'.
        \item $T=$conclusion: defined by a pair of expressions $(\Delta_1, \Delta_2)$
              and written as '$\Delta_1 = \Delta_2$'.
        \item $\in=$conclusion: defined by a triple of expressions $(t_1, t_2, \Delta)$
              and written as '$t_1 = t_2 : \Delta$'
    \end{enumerate}
\end{definition}
Finally a \textbf{rule} is determined by a premise $P$ and a
conclusion $C$, and is written as $\dfrac{P}{C}$.
We name the rule as $T$-rule, $\in$-rule, $T=$rule or $\in=$rule
according to the form of the conclusion.

Now we can define pretheory and later theory.

\begin{definition}
    A \textbf{pretheory} consists of
    \begin{enumerate}
        \item a set of sort symbols $S$;
        \item a set of operator symbols $Z$;
        \item for each sort symbol $A$, an associated rule 
              of the alphabet $S \cup Z$:
              \begin{equation*}
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    A(x_1, \dots, A_n) \text{ is a type}
                }
              \end{equation*}
              for some $n \geq 0$. It is called the introductory
              rule of $A$;
        \item for each operator symbol $F$, an associated rule
              of the alphabet $S \cup Z$:
              \begin{equation*}
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    F(x_1, \dots, x_n) : \Delta
                }
              \end{equation*}
              for some $n \geq 0$. It is called the introductory
              rule of $F$;
        \item a set of axioms. Each axiom is either a $T=$rule or
              a $\in=$rule of the alphabet $S \cup Z$.

    \end{enumerate}
\end{definition}

Now we will tackle the problem of well-formedness of rules.

\begin{definition}
    If $U$ is a pretheory, then
    \begin{enumerate}
        \item a \textbf{context} is a premise $x_1 : \Delta_1, \dots, x_n : \Delta_n$
            such that the rule
            \begin{equation*}
                \dfrac{
                    x_1 : \Delta_1, \dots, x_{n-1}: \Delta_{n-1}
                }{
                    \Delta_n \text{ is a type}
                }
            \end{equation*}
            is a derived rule such that $x_n$ is distinct from all of $x_1, \dots, x_n$.
        \item \begin{enumerate}
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  is \textbf{well-formed} if the premise $x_1 : \Delta_1, \dots, x_n : \Delta_n$ is a context.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t : \Delta}$
                  is \textbf{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  is a derived rule of $U$.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta = \Delta'}$
                  is \textbf{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  and \\ $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta' \text{ is a type}}$
                  are derived rules of $U$.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t = t' : \Delta}$
                  is \textbf{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t : \Delta}$
                  and \\ $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t' : \Delta}$
                  are derived rules of $U$.
        \end{enumerate}
    \end{enumerate}
\end{definition}

\begin{definition}
    The set of \textbf{derived rules} of a pretheory $U$ is the
    set of rules derivable by the following principles of derivation:
    
    % TODO: add rules of derivation
\end{definition}

\begin{definition}
    A pretheory is \text{well-formed} if all of its introductory
    rules are well-formed.
    A \textbf{theory} is a well-formed pretheory.
\end{definition}

\section{The Substitution Lemma}

Given $\Delta, t_1, \dots, t_n$ expressions and $x_1, \dots, x_n$ variables,
then we define the expression $\Delta[t_1 | x_1, \dots, t_n | x_n]$
the result of simultaneous replacement of $x_1, \dots, x_n$ in $\Delta$
by $t_1, \dots, t_n$. We call this the Substitution operation.

The substitution lemma states that the set of derived rules of 
a theory is closed under the operation of substitution of
correctly typed terms for variables. 
We could add substitution as one of the derivation rules
but would make applying induction very messy.

\begin{definition}
    A derived rule of $\dfrac{y_1 | \Omega_1, \dots, y_n | \Omega_n}{Conclusion}$
    of a theory $U$ is said to have the \textbf{substitution property}
    if for all contexts $Q$ of $U$, for all expressions $s_1, \dots, s_m$ of $U$
    if the rules
    \begin{equation*}
        \dfrac{Q}{s_1 : \Omega_1},
        \dfrac{Q}{s_2 : \Omega_2[s_1 | y_1]},
        \dots,
        \dfrac{Q}{s_m : \Omega_m[s_1 | y_1, \dots, s_{m-1} | y_{m-1}]}
    \end{equation*}
    are all derived rules of $U$, then the rule 
    \begin{equation*}
        \dfrac{Q}{Conclusion[s_1 | y_1, \dots, s_m | y_m]}
    \end{equation*}
    is a derived rule of $U$.
\end{definition}

We will see that all derived rules hold this property.

\begin{lemma}
    If $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{Conclusion}$
    is a derived rule of $U$, then any variable appearing in the conclusion
    occurs among $\{ x_1, \dots, x_n \}$.
\end{lemma}
\begin{proof}
    Easily obtained by applying induction over the derivation rules.
\end{proof}

\begin{lemma}
    \begin{enumerate}
        \item the premise of a derived rule is a context,
        \item if $x_1 : \Delta_1, \dots, x_n : \Delta_n$ is a context,
            then for all $1 \leq i \leq n$, then \\ 
            $\dfrac{x_1 : \Delta_1, \dots, x_{i-1} : \Delta_{i-1}}{\Delta_i \text{ is a type}}$
            is a derived rule.
    \end{enumerate}
\end{lemma}
\begin{proof}
    % TODO: label derivation rules and complete proof.
\end{proof}

\begin{lemma}[The Substitution Lemma]
    Every derived rule of a theory $U$ has the substitution property.
\end{lemma}
\begin{proof}
    % TODO: label derivation rules and complete proof.
\end{proof}

\begin{corollary}[Change of Variables]
    If $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{Conclusion}$
    is a derived rule of $U$ and if $y_1, \dots, y_n$ is a sequence of distinct variables,
    then
    \begin{equation*}
        \dfrac{
            y_1 : \Delta_1, y_2 : \Delta_2[y_1 | x_1], \dots, y_n : \Delta_n[y_1 | x_1, \dots, y_{n-1} | x_{n-1}]
        }{
            Conclusion[y_1 | x_1, \dots, y_n | x_n]
        }
    \end{equation*}
    is a derived rule of $U$.
\end{corollary}

\section{Models and Homomorphisms}

In this section we will consider the semantics of a theory.
To do this we will introduce the notion of model model
homomorphism.

First, we will set some notation and basic concepts. 
We will consider a \textbf{tree} to be a tree with a unique
least element, in the case that our tree does not have one
we will add it. This element will be denoted as $1$.
Let $\Theta$ be a tree. To denote that $A$ is a node of 
$\Theta$ we will write $A \in \Theta$, confusing the 
tree with its set of nodes. To assert that
$B$ is a node of $\Theta$ and $B$ succeeds $A$, we 
write $A \vartriangleleft B$ in $\Theta$.
Also, if $A$ is a node of $\Theta$ and $A \neq 1$,
there exists a unique $n \geq 0$ and $A_1, \dots, A_n$
such that $1 \vartriangleleft A_1 \vartriangleleft \dots \vartriangleleft A_n \vartriangleleft A$ in $\Theta$.

The set of contexts of a theory $U$ is structured as a tree.
Where its least element is the empty context and for any $n \geq 1$
the predecessor of $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$
is $\langle x_1 : \Delta_1, \dots, x_{n-1} : \Delta_{n-1} \rangle$.

We want to consider the tree of sets, families of sets, families of families of sets, \dots
Consider the following notation for families:
given $A$ is a set, if for every element $a \in A$ $B(a)$ is a set,
then the corresponding $A$-indexed family of sets is denoted as
$\lambda a \in A . B(a)$ or as $\lambda a . B(a)$;
If we have a set $C(a, b)$ for every $a \in A$ and $b \in B(a)$,
then $\lambda a \in A. \lambda b \in B(a). C(a, b)$
is an $A$-indexed family of families of sets; and so on.
The collection of sets, families of sets, families of families of sets, and so on,
is called the \textbf{tree of families}.
Given $1 \vartriangleleft A_1 \vartriangleleft A_2 \vartriangleleft \dots \vartriangleleft A_n \vartriangleleft A$
in the tree of families. Then, for all $a_1 \in A_1$, \dots,
for all $a_n \in A_n(a_1, \dots, a_{n-1})$, $A(a_1, \dots, a_n)$ is a set.
Finally we should define the concept of \textbf{operator}.
If we have $1 \vartriangleleft A_1 \vartriangleleft \dots \vartriangleleft A_n \vartriangleleft A$
in the tree of families. If for all $a_1 \in A_1$, \dots, 
for all $a_n \in A_n(a_1, \dots, a_{n-1})$,
$f(a_1, \dots, a_n) \in A(a_1, \dots, a_n)$. 
We say that $\lambda a_1 \in A_1. \dots \lambda a_n \in A_n(a_1, \dots, a_{n-1}). f(a_1, \dots, a_n)$
is an operator at $A$.

Now we will introduce the notion of a model of a theory.
It is important to understand that the derived rules are the ones
to be interpreted by the model, not expressions. 
To denote the interpretation of a rule $R$ by a model $M$
we will use the superscripted $R^M$.

\begin{definition}
    Let $U$ be a theory. $M$ is a model of $U$ if 
    \begin{enumerate}
        \item for each derived $T$-rule $R$ of the form 
            \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    \Delta \text{ is a type}
                }
            \), if we consider the rules $R_i$ to be
            \( 
                \dfrac{
                    x_1 : \Delta_1, \dots, x_{i-1} : \Delta_{i-1}
                }{
                    \Delta_i \text{ is a type}
                }
            \) for all $1 \leq i \leq n$.
            Then $M$ assigns those rules to 
            $1 \vartriangleleft R_1^M \vartriangleleft R_2^M \vartriangleleft \dots \vartriangleleft R_n^M \vartriangleleft R^M$
            in the tree of families;
        \item if $R_t$ is a derived $\in$-rule of the form 
            \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    t : \Delta 
                }
            \), following the notation of the previous point,
            $M$ assigns the rule to $R_t^M$ an operator at $R^M$.
    \end{enumerate}
\end{definition}

\begin{definition}
    If $M$ and $M'$ are models of $U$, $f : M \to M'$
    is a \textbf{homomorphism} if
    \begin{enumerate}
        \item for every derived $T$-rule $R$ of the form \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    \Delta \text{ is a type}
                }
            \) there is an operator $f_R$ such that
            for every $a_1 \in R_1^M$, \dots, for every $a_n \in R_n^M(a_1, \dots, a_{n-1})$
            and for every $a \in R^M(a_1, \dots, a_n)$,
            \[
                f_R(a_1, \dots, a_n, R_t^M(a_1, \dots, a_n)) \in
                R^{M'}(f_{R_1}(a_1), \dots, f_{R_n}(a_1, \dots, a_n))
            \]
        \item for all derived $\in$-rules $R_t$ of the form \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    t : \Delta
                }
            \), for all $a_1 \in R_1^M, \dots, a_n \in R_n^M(a_1, \dots, a_{n-1})$,
            \begin{equation*}
                f_R(a_1, \dots, a_n, R_t^M(a_1, \dots, a_n)) =
                R_t^{M'}(f_{R_1}(a_1), \dots, f_{R_n}(a_1, \dots, a_n))
            \end{equation*}
        \item if \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{
                    \Delta = \Delta'
                }
            \) is a derived rule of $U$, then $f_R = f_{R'}$,
            where $R$ and $R'$ are the rules \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{\Delta \text{ is a type}}
            \) and \(
                \dfrac{
                    x_1 : \Delta_1, \dots, x_n : \Delta_n
                }{\Delta' \text{ is a type}}
            \) respectively.
    \end{enumerate}
\end{definition}

To define a homomorphism it is enough to give a family 
$\lambda A \in S. f_A$ of correctly typed operators, where
$S$ is the set of sort symbols of $U$.

The category of models and homomorphisms of a theory $U$ is named $U-alg$.

We want to create an equivalence relation on the derived $T$- and $\in$-rules
of a theory $U$ which we call the \textbf{equivalence relation of identity of denotation}
such that $R \equiv R'$ by this relation if any model should 
interpret $R$ and $R'$ as identical objects.

\begin{example}
    Semantically we should not distinguish between the rules

    \begin{minipage}{0.4\textwidth} % TODO: fix alignment
    \begin{equation*}
        \dfrac{
            x_1 : \Delta_1, \dots, x_n : \Delta_n
        }{
            \Delta \text{ is a type}
        }
    \end{equation*}
    \end{minipage}
    \begin{minipage}{0.6\textwidth}
    \begin{equation*}
        \dfrac{
            x'_1 : \Delta_1, \dots, x'_n : \Delta_n[x'_1 | x_1, \dots, x'_{n-1} | x_{n-1}]
        }{
            \Delta[x'_1 | x_1, \dots, x'_n | x_n] \text{ is a type}
        }
    \end{equation*}
    \end{minipage}
    that only differ on the choice of the variables.
\end{example}

\begin{example}
    We should not distinguish either between 

    \begin{minipage}{0.5\textwidth} % TODO: fix alignment
    \begin{equation*}
        \dfrac{
            x_1 : \Delta_1, \dots, x_n : \Delta_n
        }{
            \Delta \text{ is a type}
        }
    \end{equation*}
    \end{minipage}
    \begin{minipage}{0.5\textwidth}
    \begin{equation*}
        \dfrac{
            x_1 : \Delta'_1, \dots, x_n : \Delta'_n
        }{
            \Delta' \text{ is a type}
        }
    \end{equation*}
    \end{minipage}
    if for each $i$, $1 \leq i \leq n$, \(
        \dfrac{
            x_1 : \Delta_1, \dots, x_{i-1} : \Delta_{i-1}
        }{
            \Delta_i = \Delta'_i
        }
    \) and \(
        \dfrac{
            x_1 : \Delta_1, \dots, x_n : \Delta_n
        }{
            \Delta = \Delta'
        }
    \).
\end{example}

We define the relation $\equiv$:
\begin{itemize}
    \item between the $T$-rules of a theory $U$ by
        \begin{equation*}
            \dfrac{
                x_1 : \Delta_1, \dots, x_n : \Delta_n
            }{
                \Delta_{n+1} \text{ is a type}
            }
            \equiv
            \dfrac{
                y_1 : \Omega_1, \dots, y_m : \Omega_m
            }{
                \Omega_{m+1} \text{ is a type}
            }
        \end{equation*}
        if, and only if, $m = n$ and for each $1 \leq i \leq n + 1$
        \begin{equation*}
            \dfrac{
                x_1 : \Delta_1, \dots, x_{i-1} : \Delta_{i-1}
            }{
                \Delta_i = \Omega_i[x_1 | y_1, \dots, x_{i-1} | y_{i-1}]
            }
        \end{equation*}
        is a derived rule of $U$.
    \item between $\in$-rules by 
        \begin{equation*}
            \dfrac{
                x_1 : \Delta_1, \dots, x_n : \Delta_n
            }{
                t : \Delta
            }
            \equiv
            \dfrac{
                y_1 : \Omega_1, \dots, y_m : \Omega_m
            }{
                s : \Omega
            }
        \end{equation*}
        if, and only if,
        \begin{equation*}
            \dfrac{
                x_1 : \Delta_1, \dots, x_n : \Delta_n
            }{
                \Delta \text{ is a type}
            }
            \equiv
            \dfrac{
                y_1 : \Omega_1, \dots, y_m : \Omega_m
            }{
                \Omega \text{ is a type}
            }
        \end{equation*}
        and \(
            \dfrac{
                x_1 : \Delta_1, \dots, x_n : \Delta_n
            }{
                t = s[x_1 | y_1, \dots, x_n | y_n] : \Delta
            }
        \) is a derived rule of $U$.
\end{itemize}

\section{Contexts and Realizations}

In this section we will introduce context categories, a powerful
interpretation of general algebraic theories that will be key 
later on. We will start by defining the category of contexts and realizations of a theory $U$.

\begin{definition}
    If $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$
    and $\langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$
    are contexts of $U$, then a \textbf{realization} of 
    $\langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$
    with regard of $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$
    is a $m$-tuple $\langle t_1, \dots, t_m \rangle$
    such that for each $j$, $1 \leq j \leq m$, the rule
    \begin{equation*}
        \dfrac{
            x_1 : \Delta_1, \dots, x_n : \Delta_n
        }{
            t_j : \Omega_j[t_1 | y_1, \dots, t_{j-1} | y_{j-1}]
        }
    \end{equation*}
    is a derived rule of $U$.
\end{definition}

Then, for a theory $U$, we define the category of contexts and realizations of $U$,
$R(U)$, as the category whose objects are contexts of $U$,
and where we have a morphism 
\begin{equation*}
    \langle t_1, \dots, t_m \rangle : 
    \langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle 
    \longrightarrow 
    \langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle
\end{equation*}
for each realization $\langle t_1, \dots, t_m \rangle$
of $\langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$
with regard of $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$.
The identity of an object $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$
is the realization $\langle x_1, \dots, x_n \rangle$.
And the composition of two morphisms
\begin{equation*}
    \langle t_1, \dots, t_m \rangle : 
    \langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle 
    \longrightarrow 
    \langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle
\end{equation*}
\begin{equation*}
    \langle s_1, \dots, s_w \rangle : 
    \langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle
    \longrightarrow
    \langle z_1 : \Lambda_1, \dots, z_w : \lambda_w \rangle
\end{equation*}
is defined as
\begin{equation*}
    \langle s_1, \dots, s_w \rangle \circ \langle t_1, \dots, t_m \rangle =
    \langle s_1[t_1 | y_1, \dots, t_m | y_m], \dots, s_w[t_1 | y_1, \dots, t_m | y_m] \rangle
\end{equation*}
 
The set of objects of $R(U)$ i.e. the set of contexts of $U$
is structured as a tree with a single least element $\langle \rangle$,
the empty context, and where every context $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n$
is preceded by $x_1 : \Delta_1, \dots, x_{n-1} : \Delta_{n-1}$.
Moreover, for any context $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n, x_{n+1} : \Delta_{n+1}, \dots, x_{n+m} : \Delta_{n+m} \rangle$
of $U$, $\langle x_1, \dots, x_n \rangle$ is a realization of
$\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$
with respect to $\langle x_1 : \Delta_1, \dots, x_{n+m} : \Delta_{n+m} \rangle$.
We name that morphism as $p(\langle x_1 : \Delta_1, \dots, x_{n+m} : \Delta_{n+m}\rangle, \langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle )$.

\begin{lemma}
    If $A = \langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$,
    $A' = \langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$ and
    $B = \langle y_1 : \Omega_1, \dots, y_{m+w} : \Omega_{m+w}$ 
    are contexts, and $f = \langle t_1, \dots, t_m \rangle$
    is a realization of $\langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$
    with regard to $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$,
    then 
    \[
    \begin{tikzcd}
        f^*B \arrow[r, "q(f {,} B)"] \arrow[d, "p(f^*B {,} A)"'] & B \arrow[d, "p(B {,} A')"] \\
        A \arrow[r, "f"'] & A'
    \end{tikzcd}
    \]
    is a pullback diagram in $R(U)$, where
    \begin{multline*}
    f^*B = \langle x_1 : \Delta_1, \dots, x_n : \Delta_n, y_{m+1} : \Omega_{m+1}[t_1 | y_1, \dots, t_m | y_m], \\
    \dots, y_{m+w} : \Omega_{m+w}[t_1 | y_1, \dots, t_m | y_m] \rangle    
    \end{multline*}
    \begin{equation*}
        q(f, B) = \langle t_1, \dots, t_m, y_{m+1}, \dots, y_{m+w} \rangle
    \end{equation*}
\end{lemma}

% TODO: should I prove this?
This properties motivate the following definition:

\begin{definition}
    A \textbf{contextual category} consists of
    \begin{enumerate}
        \item a category $C$ with a terminal object $1$;
        \item a tree structure on the objects of $C$ such that
              the terminal object $1$ is the unique least element
              of the tree;
        \item for all $A, B \in Ob C$ such that $A \vartriangleleft B$
              a morphism $p(B) : B \to A$ in $C$ and that will be
              written as $B \openrightarrow A$;
        \item for all $A, A' \in Ob C$ and for all $f: A \to A'$ in $C$,
              for all $B \in Ob C$ such that $A' \vartriangleleft B$,
              an object $f^*B$ in $C$ and a morphism $q(f, B): f^*B \to B$
              such that the diagram
              \[
                \begin{tikzcd}
                    f^*B \arrow[r, "q(f {,} B)"] \arrow[d, open] & B \arrow[d, open] \\
                    A \arrow[r, "f"'] & A'
                \end{tikzcd}
              \]
              is a pullback in $C$ such that
              \begin{enumerate}
                \item for all $A, B \in Ob C$ such that $A \vartriangleleft B$,
                      $id_A^*B = B$ and $q(id_A, B) = id_B$;
                \item when
                    \[
                    \begin{tikzcd}
                        && B \arrow[d, open] \\
                        A \arrow[r, "f"] & A' \arrow[r, "f'"] & A''
                    \end{tikzcd}
                    \]
                    in $C$, then $(f' \circ f)^*B = f^*(f'^*B)$
                    and $q(f' \circ f, B) = q(f', B) \circ q(f, f'^*B)$.
              \end{enumerate}
    \end{enumerate}
\end{definition}

The category $R(U)$ is structured as a contextual category by the 
tree structure and pullbacks defined earlier.

The contextual category $R(U)$ has an homomorphic image $C(U)$
induced by the equivalence relation of intended equivalence of denotation.
More precisely, $C(U)$ is the category of equivalence classes
of contexts and realizations under the following relation:
\begin{itemize}
    \item Among contexts:
        \begin{equation*}
            \langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle
            \equiv
            \langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle
        \end{equation*}
        if, and only if,
        \begin{equation*}
            \dfrac{
                x_1 : \Delta_1, \dots, x_{n-1} : \Delta_{n-1}
            }{
                \Delta_n \text{ is a type}
            }
            \equiv
            \dfrac{
                y_1 : \Omega_1, \dots, y_{m-1} : \Omega_{m-1}
            }{
                \Omega_m \text{ is a type}
            }
        \end{equation*}
    \item If we have two realizations 
    $\lambda t_1, \dots, t_m \rangle$ of $\langle y_1 : \Omega_1, \dots, y_m : \Omega_m \rangle$
    with regard to $\langle x_1 : \Delta_1, \dots, x_n : \Delta_n \rangle$, and
    $\lambda s_1, \dots, s_m \rangle$ of $\langle y'_1 : \Omega'_1, \dots, y'_m : \Omega'_m \rangle$
    with regard to $\langle x'_1 : \Delta'_1, \dots, x'_n : \Omega'_n \rangle$,
    we say that $\langle t_1, \dots, t_m \rangle \equiv \langle s_1, \dots, s_m \rangle$
    if, and only if, for each $1 \leq j \leq m$,
    \begin{equation*}
        \dfrac{
            x_1 : \Delta_1, \dots, x_n : \Delta_n
        }{
            t_j : \Omega_j[t_1 | y_1, \dots, t_{j-1} | y_{j-1}]
        }
        \equiv
        \dfrac{
            x'_1 : \Delta'_1, \dots, x'_n : \Delta'_n
        }{
            s_j : \Omega'_j[s_1 | y'_1, \dots, s_{j-1} | y'_{j-1}]
        }
    \end{equation*}
\end{itemize}

We will define $Fam$, a category that plays the same role
among contextual categories as $Set$ among categories. 
The tree of objects of $Fam$ is the tree of sets, families 
of sets, families of families of sets and so on, with a formally
adjoined least element $1$.
For $n, m \geq 0$, 
if $1 \vartriangleleft A_1 \vartriangleleft \dots \vartriangleleft A_n$
and $1 \vartriangleleft B_1 \vartriangleleft \dots \vartriangleleft B_m$
in $Fam$, then the set of morphisms of $A_n$ into $B_n$,
is the set of $m$-tuples $\langle f_1, \dots, f_m \rangle$
where $f_1, \dots, f_m$ are operators such that for each $1 \leq j \leq m$,
for $a_1 : A_1, \dots, a_n : A_n(a_1, \dots, a_{n-1})$,
$f_j(a_1, \dots, a_n) \in B_j(f_1(a_1, \dots, a_n), \dots, f_{j-1}(a_1, \dots, a_n))$.
For $n = 0$, $Hom_{Fam}(1, B_m) = \{ 
    \langle b_1, \dots, b_m \rangle | b_1 \in B_1, \dots, b_m \in B_m(b_1, \dots, b_{m-1})
\}$. And for $m = 0$, $Hom_{Fam}(A_n, 1) = \{ \langle \rangle \}$,
hence $1$ is the terminal object of $Fam$.
To define composition in $Fam$ let 
$1 \vartriangleleft A_1 \vartriangleleft \dots \vartriangleleft A_n$,
$1 \vartriangleleft B_1 \vartriangleleft \dots \vartriangleleft B_m$,
$1 \vartriangleleft C_1 \vartriangleleft \dots \vartriangleleft C_w$,
$\langle f_1, \dots, f_m \rangle : A_n \to B_m$ and 
$\langle g_1, \dots, g_w \rangle : B_m \to C_w$, then 
$\langle g_1, \dots, g_w \rangle 
\circ \langle f_1, \dots, f_m \rangle 
= \langle h_1, \dots, h_w \rangle$ where for all
$1 \leq k \leq w$, $h_k$ is defined as
\begin{equation*}
    h_k(a_1, \dots, a_n) = g_k(f_1(a_1, \dots, a_n), \dots, f_m(a_1, \dots, a_n))
\end{equation*}
% TODO: Continue