% !TEX root = ../main.tex
\addcontentsline{toc}{chapter}{Introduction}
\chapter*{Introduction}
% TODO: Abstract appears in the header!!

Our ultimate goal in this thesis is to translate the syntactical problem of theorem proving into a geometrical graph-exploration problem.

The idea of generating mathematical reasoning computationally
is one of the most profound challenges in computer science.
At the core of this pursuit lies the Curry-Howard correspondence,
an isomorphism between logic and computation \cite{howard}:
mathematical propositions are interpreted as types, and the proofs
of propositions are viewed as terms (or programs) inhabiting
those types. Consequently, the problem of proving a theorem is 
reduced to the problem of finding a well-typed term in a formal type theory.

Modern proof assistants, such as Rocq (formerly Coq) \cite{huet2000} and Lean \cite{demoura2015}
leverage this correspondence using rich type theories like the 
Calculus of Inductive Constructions \cite{coquand1990,paulin-mohring1993}. While these environments
enabled the formalization of highly complex mathematics, fully autonomous theorem
proving with such rich frameworks remains an extremely
difficult challenge. The primary obstacle is undecidability of 
type inhabitation in dependent type theories \cite{dowek1993}, and even for inhabited types, attempting
to exhaustively construct a term of a given type results in a 
combinatorial explosion of the search space.

Traditionally, term generation has been approached from a purely
syntactic perspective, and more recently through Large Language Models (LLMs).
Generating terms in Minimal Implicational Combinatory Logic using LLMs
was the starting point of our research, a problem proposed by Dr. Joost J. Joosten.
However, generating terms from syntax alone is difficult to guide heuristically,
and the linear nature of language model generation forces early commitments 
that can prevent a proof from ever being found. Similar limitations 
have been observed by the Logical Intelligence team in \cite{logicalintelligence}.

We were working with Topos Theory \cite{lawvere1971} at the time, in an attempt 
to understand the framework proposed by Laurent Lafforgue \cite{lafforgue2022,lafforgue2024}, which connects
Topos Theory with Artificial Intelligence. 
It became clear that we needed to leverage the categorical semantics 
of type theories to reframe autonomous theorem proving as a 
structural graph exploration problem.
Then we developed the \textsc{implica} graph engine together with the first version of the 
expansion algorithm.

Then with the aim of expanding outside Minimal Implicational Combinatory Logic, we 
started our research about Generalized Algebraic Theories (GATs) \cite{cartmell1978,cartmell1986} and the Calculus of Constructions \cite{coquand1988}.
They both offered a way of translating the syntax into categorical structures 
---namely Contextual Categories and Doctrines of Constructions \cite{streicher}--- 
to obtain term models that carry a rich geometric graph structure. 
The soundness of this translation is guaranteed by the term 
model being the initial object of the category of models of 
the underlying type theory, this is known as the Initiality Conjecture. 
The initiality of the term model for GATs was originally established by Cartmell \cite{cartmell1978,cartmell1986},
And the initiality of the term model for CoC was established by Streicher \cite{streicher}.
The modern Initiality Conjecture, championed by Voevodsky \cite{voevodsky2017}, pertains to the much more complex Martin-L\"of Type Theory, 
which remains beyond the scope of our current engine, and that was proven by Brunerie and Lumsdaine \cite{bruneire2020}.

In this categorical framework, the problem of finding a proof
translates to finding a specific section ---a specific morphism--- 
within a relativized contextual category. 
With this idea in mind of using categories and neural networks, we formalized 
an expansion algorithm 
---a deterministic process that exhaustively builds the finite subcategories
representing the search space of terms--- specifically designed to work together
with Graph Neural Networks (GNNs), and try to overcome the severe exponential growth
inherent to any algorithm trying to exhaustively solve the problem.

Graph Neural Networks are used to navigate and prune the categorical search space. 
Considering our finite subcategories that form
the known search space as graphs, they are the perfect inputs 
for message-passing neural networks. By training a GNN to classify
nodes in the context graph (deciding which paths to expand, keep
or delete), we can effectively guide the expansion algorithm,
bypassing the combinatorial explosion and paving the way for efficient,
autonomous theorem proving.

\section*{Objectives}

We provide an explicit geometric formulation of Cartmell's term models, adapting them to be computationally navigable by a GNN.
Building on this, we extend the Calculus of Constructions to accommodate a GAT signature, 
define the corresponding term model, and outline how Streicher's initiality proof generalizes to this richer framework.
Then we introduce the Expansion Algorithm, a deterministic procedure for generating terms within the term models described above, 
specifically designed to interface with a Graph Neural Network so as to guide the search and counteract the combinatorial 
explosion inherent to exhaustive term construction. 
To ground this algorithm and build intuition, we provide a full implementation for Simply Typed Combinatory Logic ($\textbf{CL}_{\to}$)
---a tractable yet representative system--- using \textsc{implica}, a custom graph engine written in Rust. 
Finally, we present a design proposal for the GNN-based model, including a novel unsupervised Conjecturer-Prover training paradigm intended to optimize 
the expansion process toward efficient, autonomous theorem proving.

\section*{Prerequisites}

This thesis bridges type theory, categorical semantics, and machine learning. 
While specialized categorical structures ---such as Contextual Categories and Doctrines of Constructions--- are introduced and defined within the text, 
a basic familiarity with the fundamental concepts of category theory (e.g., categories, functors, and pullbacks) is assumed. 
For readers seeking an introduction or a refresher on this foundational material, Awodey's \textit{Category Theory} \cite{awodey2010} is an excellent and highly recommended reference, 
particularly for its focus on logic and computer science. 
As a more classical reference there is Mac Lane's \textit{Categories for the Working Mathematician} \cite{maclane1992}.

\section*{Thesis Outline}
This thesis is structured to guide the reader from the abstract categorical foundations
to the practical implementation of AI-guided theorem proving:
\begin{itemize}
    \item \textbf{Chapter 1: Syntax and Semantics of Generalized Algebraic Theories.}
        We introduce Cartmell's Generalized Algebraic Theories and define their syntax and semantics, as 
        they were originally presented in \cite{cartmell1978, cartmell1986} and incorporate some notions
        presented by Streicher in \cite{streicher}.
        Then we provide a concept of model of a GAT together with a proof of the initiality of the term model.
        Finally, we define the problem of type inhabitation in categorical terms, and present the foundational Expansion Algorithm.
    \item \textbf{Chapter 2: Calculus of Constructions.} 
        We present the syntactical definition of CoC together with the construction 
        of Doctrines of Constructions as models for CoC, given by Streicher in \cite{streicher}.
        We extend the definition of CoC to include a GAT signature, define the new term model and outline a proof of its initiality.
        Finally, we extend our categorical problem definition and 
        Expansion Algorithm to handle this richer syntax.
    \item \textbf{Chapter 3: Implementation of the Expansion Algorithm.} 
        We demonstrate the raw Expansion Algorithm generating terms of Simply Typed Combinatory 
        Logic ($\textbf{CL}_\to$) using \textsc{implica}, a custom Rust-based graph engine written by the author (consists of $\sim 7.000$ lines of Rust code). 
        Finally, we outline the architecture for integrating Graph Neural Networks
        using a novel unsupervised `Conjecturer-Prover' training paradigm to optimize the expansion process.
\end{itemize}

