% !TEX root = ../main.tex
\chapter{The Syntax and Semantics of Generalized Algebraic Theories}

In this chapter, we introduce Generalized Algebraic Theories (GATs), a framework for working with 
dependent types. GATs allow us to construct the concept of \textit{contextual category},
the foundational categorical structure to deal with dependent types. We use contextual categories as a baseline 
for interpreting the more complex Calculus of Constructions in Chapter 2.

\section{Generalized Algebraic Theories}

Generalized Algebraic Theories, originally presented by J. Cartmell 
in his thesis \cite{cartmell1978} \cite{cartmell1986},
is an abstraction of Martin-L\"of type theory
and is a generalization of the usual notion of a many-sorted 
algebraic theory in the following manner:
while in Martin L{\"o}f's many-sorted algebraic theories sorts
are constant types, interpreted as sets, in generalized
algebraic theories sorts might be constant or variable,
over a specified range of types. Variable types
are interpreted as families of sets. 
We will start by giving some intuition, and we will later give the formal definition.

Informally, a generalized
algebraic theory consists of:
\begin{enumerate}
    \item a set of sorts, that can be constant or variable;
    \item a set of operators, with the sorts of the arguments and value specified;
    \item a set of axioms, where each axiom is an identity between similar expressions. 
\end{enumerate}

To build intuition, consider the theory of categories.
The set of sorts contains two elements $\Ob$ and $\Hom$.
$\Ob$ is a constant type and $\Hom$ is a variable type;
for any two terms $x, y$ of type $\Ob$, we have a type $\Hom(x,y)$.
Here, we see a key feature of GATs: types vary over terms, not over types.
The set of operators also has two elements $\id$ and $\circ$.
The $\id$ operator has one argument $x$ of type $\Ob$,
and returns a value of type $\Hom(x, x)$. This shows
how the type of value of an operator might vary depending
on the arguments. 
The $\circ$ operator (or composition) expects 5 arguments, 
three terms $x, y, z$ of type $\Ob$, one term $f$ of type $\Hom(x, y)$,
and one term $g$ of type $\Hom(y, z)$. It returns
a value of type $\Hom(x, z)$.
Here we see that the type of an argument might also vary depending
on a previous argument.
Informally, instead of writing $\circ(x, y, z, f, g)$ we can
write $\circ(f, g)$, as the first three arguments might be inferred from the last two.
This notation is discussed in the original paper \cite{cartmell1986}. 

Now we introduce the notation we will be using when talking 
about terms and types. This notation will be formally defined
in a later section. Given a set of variables $V$,
we associate types to variables as required.
When asserting that a term $t$ is of type $A$, we will write
$t \colon A$, but if $t$ has variables $x_1, \dots, x_n$ occurring in it, this assertion
does not make sense unless we assign some types to $x_1, \dots, x_m$.
To do this we write $x_1 \colon A_1, \dots, x_n \colon A_n \vdash t \colon A$ or
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n: A_n
    }{
        t : A
    }
\end{equation*}
that reads as follows: if $x_1$ is a variable of type $A_1$, \dots and if $x_n$ is a variable of type $A_n$,
then $t$ is a term of type $A$. 
Similarly, if we want to assert that if $x_1$ is a variable of type $A_1$,
\dots and if $x_n$ is a variable of type $A_n$,
then $A$ is a type, we can write: $x_1 \colon A_1, \dots, x_n \colon A_n \vdash A \type$ or
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n 
    }{
        A \type
    }
\end{equation*}
Such assertions are called rules or sequents. 
Instead of considering expressions or judgments as basic units, we will use rules, 
as both expressions and judgments make sense only within a context.

Axioms are also written as rules. The conclusion of those rules
are equalities between types or between terms of the same type.
We write those rules as follows:

\begin{minipage}{0.5\textwidth}
  \begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        A = A'
    }
  \end{equation*}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
  \begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        t = t' : A
    }
  \end{equation*}
\end{minipage}

and read as: if $x_1$ is a variable of type $A_1$, \dots and $x_n$ is a variable of type $A_n$,
then $A = A'$, or $t = t'$ as terms of type $A$, respectively.
We will be using both shorthand and fraction notation as needed.

% TODO: add comments on x, y: A <-> x : A, y : A

% TODO: fix alignment
To continue with the previous example, we can write the axioms
of category theory as follows:
\begin{alignat*}{2}
    &x, y : Ob, f : Hom(x, y)  \vdash  \circ(id(x), f) = f, \\
    &x, y : Ob, f : Hom(x, y)  \vdash  \circ(f, id(y)) = f, \\
    &w, x, y, z : Ob, f : Hom(w, x), g : Hom(x, y), h : Hom(y, z) \\ 
        &\qquad\vdash \circ(\circ(f, g), h) = \circ(f, \circ(g, h)).
\end{alignat*}

When presenting a theory's language i.e., the sets of sorts and operators,
we must present each symbol with a rule specifying the role
of this symbol, either as a sort symbol or as specifically typed operator symbol. 
These rules are called \textit{introductory rules}.
In the case of a sort symbol $A$, the rule should be like 
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        A(x_1, \dots, x_n) \type
    }
\end{equation*}
clearly specifying the types $A$ is dependent on.
In the case of an operator symbol, a rule of the form
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        f(x_1, \dots, x_n) : A
    }
\end{equation*}
suffices to specify the types of the arguments and the type of the value of $f$.

Finally, to present a theory we will give a set of symbols,
each associated to its introductory rule, and a set of axioms.
And, of course, everything must be well-formed, {\color{orange} something we 
will address in the following section.}
Now we give a formal presentation of category theory using GAT notation: 
in Table \ref{table:categories-symbols} we state the symbols of the theory together with their introductory rules
and in Table \ref{table:categories-axioms} we state the axioms of the theory as rules.

% --- Table 1: Introductory Rules ---
\begin{table}[htbp]
  \centering
  \renewcommand{\arraystretch}{1.4}
  \begin{tabular}{@{} l p{10cm} @{}}
    \toprule
    \textit{Symbol} & \textit{Introductory Rule} \\
    \midrule
    $\Ob$   & $\vdash \Ob  \type$ \\
    $\Hom$  & $x, y \colon \mathrm{Ob} \vdash \Hom(x, y) \type$ \\
    $\circ$         & $x, y, z \colon \Ob, f \colon \Hom(x,y), g \colon \Hom(y,z) \vdash \circ(f,g) \colon \Hom(x,z)$ \\
    $\mathrm{id}$   & $x \in \Ob \vdash \id(x) \colon \Hom(x,x)$ \\
    \bottomrule
  \end{tabular}
  \caption{Symbols and their introductory rules.}
  \label{table:categories-symbols}
\end{table}

% --- Table 2: Axioms ---
\begin{table}[htbp]
  \centering
  \renewcommand{\arraystretch}{1.6}
  \begin{tabular}{@{} p{6.5cm} p{8cm} @{}}
    \toprule
    \multicolumn{2}{@{}l}{\textit{Axioms}} \\
    \midrule
    $\circ(\id(x), f) = f$
      & whenever $x, y \colon \Ob$ and $f \colon \Hom(x, y)$ \\
    $\circ(f, \id(y)) = f$
      & whenever $x, y \colon \Ob$ and $f \colon \Hom(x, y)$ \\
    $\circ(\circ(f,g),\,h) = \circ(f,\circ(g,h))$
      & whenever $w, x, y, z \colon \Ob$, $f \colon \Hom(w,x)$,
        $g \colon \Hom(x,y)$ and $h \colon \Hom(y,z)$ \\
    \bottomrule
  \end{tabular}
  \caption{Axioms (left unit, right unit, associativity).}
  \label{table:categories-axioms}
\end{table}

\section{Formal Definition}

In this section we formally define all of the concepts 
we just introduced.
To give a notion of well-formed introductory rule, we need the concept of derivability,
which also depends on the introductory rules. 
To tackle this problem, we leave the question of
well-formedness aside until we have the complete set of derived rules
of a theory.
We first define a pre-theory and then define a theory
as a well-formed pre-theory.

We assume given a countable set of variables $V$.
Let us define the notion of rule on an alphabet $W$,
whose elements are called symbols.

\begin{definition}
    The set of \textit{expressions} $\mathcal{E}$, is defined inductively
    so that expressions are finite sequences of elements
    of $W \cup V \cup \{(\} \cup \{)\} \cup \{,\}$
    defined by the following clauses:
    \begin{enumerate}
        \item if $x \in V$, then $x$ is an expression;
        \item if $L \in W$, then $L$ is an expression;
        \item if $L \in W$ and $e_1, \dots, e_n$ are expressions,
              then $L(e_1, \dots, e_n)$ is an expression.
    \end{enumerate}
\end{definition}

Now we can define the building blocks of a rule:
\begin{definition}
    A \textit{premise} is defined to be a sequence of elements of $V \times \mathcal{E}$.
    The empty sequence is allowed and we call it an empty premise.
    If $((x_1, A_1), \dots, (x_n, A_n))$ is a premise,
    we write it as $x_1 : A_1, \dots, x_n : A_n$.
\end{definition}
\begin{definition}
    We call a \textit{conclusion} to one of the following:
    \begin{enumerate}
        \item $T$-conclusion: defined by a single expression $A$
              and written as '$A \type$'.
        \item $\in$-conclusion: defined by a pair of expressions $(t, A)$
              and written as '$t \colon A$'.
        \item $T=$conclusion: defined by a pair of expressions $(A_1, A_2)$
              and written as '$A_1 = A_2$'.
        \item $\in=$conclusion: defined by a triple of expressions $(t_1, t_2, A)$
              and written as '$t_1 = t_2 : A$'
    \end{enumerate}
    Finally a \textit{rule} is determined by a premise $P$ and a
    conclusion $C$, and is written as $\dfrac{P}{C}$ or as $P \vdash C$.
    We name the rule as $T$-rule, $\in$-rule, $T=$rule or $\in=$rule
    according to the form of the conclusion.
\end{definition}

Now we can define a \textit{pre-theory} and later a \textit{theory}.

\begin{definition}
    A \textit{pre-theory} consists of
    \begin{enumerate}
        \item a set of sort symbols $\mathcal{S}$;
        \item a set of operator symbols $\mathcal{Z}$;
        \item for each sort symbol $A \in \mathcal{S}$, an associated rule 
              of the alphabet $\mathcal{S} \cup \mathcal{Z}$:
              \begin{equation*}
                \dfrac{
                    x_1 : A_1, \dots, x_n : A_n
                }{
                    A(x_1, \dots, x_n) \type
                }
              \end{equation*}
              for some $n \geq 0$. It is called the \textit{introductory
              rule} of $A$;
        \item for each operator symbol $F \in \mathcal{Z}$, an associated rule
              of the alphabet $\mathcal{S} \cup \mathcal{Z}$:
              \begin{equation*}
                \dfrac{
                    x_1 : A_1, \dots, x_n : A_n
                }{
                    F(x_1, \dots, x_n) : A
                }
              \end{equation*}
              for some $n \geq 0$. It is called the \textit{introductory
              rule} of $F$;
        \item a set of axioms. Where each axiom is either a $T=$rule or
              a $\in=$rule of the alphabet $S \cup Z$.

    \end{enumerate}
\end{definition}

\begin{definition}
    A pre-theory is \textit{well-formed} if all of its introductory
    rules are well-formed.
    A \textit{theory} is a well-formed pre-theory.
\end{definition}

Now we tackle the problem of well-formedness of rules.

\begin{definition}
    If $U$ is a pre-theory, then
    \begin{enumerate}
        \item a \textit{context} is a premise $x_1 : A_1, \dots, x_n : A_n$
            such that the rule
            \begin{equation*}
                \dfrac{
                    x_1 : A_1, \dots, x_{n-1}: A_{n-1}
                }{
                    A_n \type
                }
            \end{equation*}
            is a derived rule such that $x_n$ is distinct from all of $x_1, \dots, x_n$.
        \item \begin{enumerate}
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  is \textit{well-formed} if the premise $x_1 : \Delta_1, \dots, x_n : \Delta_n$ is a context.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t : \Delta}$
                  is \textit{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  is a derived rule of $U$.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta = \Delta'}$
                  is \textit{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta \text{ is a type}}$
                  and \\ $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{\Delta' \text{ is a type}}$
                  are derived rules of $U$.
            \item the rule $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t = t' : \Delta}$
                  is \textit{well-formed} if $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t : \Delta}$
                  and \\ $\dfrac{x_1 : \Delta_1, \dots, x_n : \Delta_n}{t' : \Delta}$
                  are derived rules of $U$.
        \end{enumerate}
    \end{enumerate}
\end{definition}

\begin{definition}
    The set of \textit{derived rules} of a pre-theory $U$ is the
    set of rules derivable by the derivation rules presented in Appendix \ref{app:derivation-rules-gat}.
\end{definition}

Lastly, we introduce \textit{substitution}.
Given expressions $\Delta, t_1, \dots, t_n$ and variables $x_1, \dots, x_n$,
then we define the expression $\Delta[x_1 \leftarrow t_1, \dots, x_n \leftarrow t_n]$
as the result of simultaneous replacement of $x_1, \dots, x_n$ in $\Delta$
by $t_1, \dots, t_n$. We call this the substitution operation.
Then, as proven in the original paper \cite{cartmell1986}, 
the set of derived rules of a theory is closed under the 
operation of substitution of correctly typed terms 
for variables, i.e., that we can substitute a variable $x$ of type $A$ in an expression
only by terms $t$ such that we can derive that $t \colon A$.
We could add substitution as one of the derivation rules,
but that would make applying induction very messy.

\section{Contexts and Realizations} % --- Notation Changes ---

To transition from syntax to geometry (which our Graph Neural Network will eventually navigate),
we construct a category \cite{maclane1971} out of the theory syntax.

\begin{definition}
    For a theory $U$, we define the \textit{category of contexts and realizations},
    denoted $R(U)$.
    This category has as objects the contexts of $U$.
    Given two contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$ and 
    $\Delta = y_1 \colon B_1, \dots, y_m \colon B_m$ of $U$, a morphism from $\Gamma$ to $\Delta$,
    or a \textit{realization} of $\Delta$ from $\Gamma$, is an $m$-tuple $\langle t_1, \dots, t_m \rangle$ 
    such that for each $j$, $1 \leq j \leq m$, the rule 
    \[
        \Gamma \vdash t_j \colon B_j[y_1  \leftarrow  t_1, \dots, y_{j-1}  \leftarrow  t_{j-1}] 
    \]
    is derivable.
    The identity of an object $x_1 \colon \Delta_1, \dots, x_n \colon \Delta_n$ is the realization 
    $\langle x_1, \dots, x_n \rangle$, and composition is defined as 
    \begin{multline*}
        \langle t_1, \dots, t_m \rangle \circ \langle s_1, \dots, s_w \rangle = \\
        \langle s_1[y_1  \leftarrow  t_1, \dots, y_m  \leftarrow  t_m], \dots, s_w[y_1  \leftarrow  t_1, \dots, y_m  \leftarrow  t_m] \rangle
    \end{multline*}
    whenever $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$ and $\Delta = y_1 \colon B_1, \dots, y_m \colon B_m$
    and $\Omega = z_1 \colon C_1, \dots,  z_w \colon C_w$ are contexts, and 
    $\langle t_1, \dots, t_m \rangle : \Gamma \to \Delta$ and $\langle s_1, \dots, s_w \rangle : \Delta \to \Omega$
    are realizations.
\end{definition}

Now we study some properties of this category.
First, we see that it has a tree structure.
\begin{definition}
    A \text{tree} is a pair $(N, \parent)$ where 
    $N$ is a nonempty set of nodes and 
    $\parent: N \to N$ is a function such that:
    \begin{enumerate}
        \item for all $A \in N$, there exists $n \in \omega$
              such that $\parent^n(A) = \parent^{n+1}(A)$;
        \item \label{item:tree-ii}
              for all $A, B \in N$, if $A = \parent(A)$ and
              $B = \parent(B)$, then $A = B$.
    \end{enumerate}
\end{definition}
We write $A \unlhd B$ if $A = \parent(B)$
and $A \lhd B$ if $A \unlhd B$ and $A \neq B$.
By Condition \ref{item:tree-ii}, there is a unique
least element of $N$, which we call $1$.
Finally, we will define, for any $A \in N$,
$\level(A)$ to be the least $n \in \omega$ with 
$\parent^n(A) = \parent^{n+1}(A)$.
We will consider $\leq$ to be the transitive completion of the relation $\unlhd$.
We can see that the set of objects of $R(U)$, i.e., the set of contexts of $U$, is structured as a tree
with the empty context as its least element and 
\(
    \parent (\langle x_1 \colon A_1, \dots, x_n \colon A_n\rangle)
    = 
    \langle x_1 \colon A_1, \dots, x_{n-1} \colon A_{n-1} \rangle
\).
Moreover, given a context $x_1 \colon A_1, \dots, x_n \colon A_n, x_{n+1} \colon A_{n+1}, \dots x_{n+m}$
of $U$, $\langle x_1, \dots, x_n \rangle$ is a realization of $x_1 \colon A_1, \dots, x_n \colon A_n$
with regard to $x_1 \colon A_1, \dots, x_n \colon A_n, x_{n+1} \colon A_{n+1}, \dots, x_{n+m} \colon A_{n+m}$, and is noted
as $p(\langle x_1 \colon A_1, \dots, x_{n+m} \colon A_{n+m} \rangle, \langle x_1 \colon A_1, \dots, x_n \colon A_n \rangle)$.
Hence, if we consider $\Gamma, \Delta \in \Ob R(U)$ such that $\Gamma \leq \Delta$, then $p(\Delta, \Gamma)$ is defined and 
$p(\Delta, \Gamma) : \Delta \to \Gamma$ in $R(U)$.

If $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$, $\Gamma' = y_1 \colon B_1, \dots, y_m \colon B_m$
and $\Delta = y_1 \colon B_1, \dots, y_{m+w} \colon B_{m+w}$ are contexts, and if 
$f = \langle t_1, \dots, t_m \rangle : \Gamma \to \Gamma'$ in $R(U)$, then
\[
\begin{tikzcd}
    f^*\Delta \arrow[r, "{q(f, \Delta)}"] \arrow[d, "{p(f^*\Delta, \Gamma)}"'] & \Delta \arrow[d, "{p(\Delta, \Gamma')}"] \\
    \Gamma \arrow[r, "f"] & \Gamma'
\end{tikzcd}
\]
is a pullback diagram \cite{maclane1971} in $R(U)$, where % TODO: fix alignment
\begin{multline*}
    f^*\Delta 
    = 
        x_1 \colon A_1, \dots, x_n \colon A_n, \\
        y_{m+1} \colon B[y_1  \leftarrow  t_1, \dots, y_m  \leftarrow  t_m], 
        \dots,
        y_{m+w} \colon B[y_1  \leftarrow  t_1, \dots, y_m  \leftarrow  t_m] 
\end{multline*}
\[
    q(f, \Delta) = \langle t_1, \dots, t_m, y_{m+1}, \dots, y_{m+w} \rangle.
\]

This motivates the following definition:

\section{Contextual Categories}

\begin{definition}
    A \textit{contextual category} consists of
    \begin{enumerate}
        \item a category $\mathcal{C}$ with a terminal object $1$;
        \item a function $\parent \colon \Ob \mathcal{C} \to \Ob \mathcal{C}$ 
              such that $(\Ob \mathcal{C}, \parent)$ is a tree, and $1$ is the root of the tree, 
              i.e., $\parent 1 = 1$;
        \item for each object $A$ of $\mathcal{C}$ different from $1$, a morphism $p(A) \colon A \to \parent A$,
              called \textit{canonical projection of $A$}, which we write as $A \openrightarrow \parent(A)$;
        \item for all objects $A, B$ in $\mathcal{C}$ with $A \lhd B$ and all morphisms $f \colon C \to A$, 
              an object $f^*B$ of $\mathcal{C}$ such that $C \lhd f^*B$, and a morphism $q(f, B) \colon f^*B \to B$
              in $\mathcal{C}$ such that
              \[
                \begin{tikzcd}
                    f^*B \arrow[r, "{q(f, B)}"] \arrow[d, open] & B \arrow[d, open] \\
                    C \arrow[r, "f"] & A
                \end{tikzcd}
              \]
              is a pullback in $\mathcal{C}$ that satisfies the following \textit{functoriality conditions}:
              \begin{enumerate}
                \item if $A$ and $B$ are objects in $\mathcal{C}$ such that $A \lhd B$, then 
                      \[
                        id_A^*B = B \quad \text{ and } \quad q(id_A, B) = id_B;
                      \]
                \item whenever 
                      \[
                        \begin{tikzcd}
                            & & B \arrow[d, open] \\
                            A \arrow[r, "f"] & A' \arrow[r, "f'"] & A''
                        \end{tikzcd}
                      \]
                      in $\mathcal{C}$, then 
                      \[
                        (f' \circ f)^*B = f^*(f'^*B) 
                            \quad \text{ and } \quad 
                        q(f' \circ f, B) = q(f', B) \circ q(f, f'^*B).
                      \]
              \end{enumerate}
    \end{enumerate}
\end{definition}

We have that $R(U)$ is structured as a contextual category. The problem with $R(U)$ is that semantically equal 
but syntactically different rules are interpreted as different elements. Hence we define an equivalence
relation $\equiv$ between derived $T$- and $\in$- rules as follows:
\[
        x_1 \colon A_1, \dots, x_n \colon A_n
        \vdash
        A_{n+1} \type
    \equiv
        y_1 \colon B_1, \dots, y_m \colon B_m
        \vdash
        B_{m+1} \type
\]
if and only if $m = n$ and, for each $1 \leq i \leq n + 1$, \[
     x_1 \colon A_1, \dots, x_{i-1} \colon A_{i-1} \vdash A_i = B_i[y_1  \leftarrow  x_1, \dots, y_{i-1}  \leftarrow  x_{i-1}]
\] is derivable,
and 
\[
        x_1 \colon A_1, \dots, x_n \colon A_n 
        \vdash 
        t : A
    \equiv
        y_1 \colon B_1, \dots, y_m \colon B_m
        \vdash
        s : B
\]
if and only if \[
        x_1 \colon A_1, \dots, x_n \colon A_n
        \vdash
        A \type
    \equiv
        y_1 \colon B_1, \dots, y_m \colon B_m
        \vdash
        B \type
\] and \[ 
    x_1 \colon A_1, \dots, x_n \colon A_n 
    \vdash t = s[y_1  \leftarrow  x_1, \dots, y_m  \leftarrow  x_m] : A
\] is derivable.

We define the \textit{context category} of $U$, denoted by $C(U)$, as the category of equivalence classes of contexts and of equivalence
classes of realizations. Two contexts $x_1 \colon A_1, \dots, x_n \colon A_n$ and 
$y_1 \colon B_1, \dots, y_m \colon B_m$ are equivalent if and only if \[
        x_1 \colon A_1, \dots, x_{n-1} \colon A_{n-1}
        \vdash
        A_{n} \type
    \equiv
        y_1 \colon B_1, \dots, y_{m-1} \colon B_{m-1}
        \vdash
        B_{m} \type
\] and two realizations 
\begin{align*}
    \langle t_1, \dots, t_m \rangle 
        & \colon \langle x_1 \colon A_1, \dots, x_n \colon A_n \rangle
         \to  \langle y_1 \colon B_1, \dots, y_m \colon B_m \rangle \\
    \langle s_1, \dots, s_m \rangle
        & \colon \langle x'_1 \colon A'_1, \dots, x'_n \colon A'_n \rangle 
         \to \langle y'_1 \colon B'_1, \dots, y'_m \colon B'_m \rangle
\end{align*}
are equivalent if and only if, for each $j$, $1 \leq j \leq n$,
\begin{multline*}
     x_1 \colon A_1, \dots, x_n \colon A_n
        \vdash
        t_j \colon B_j[y_1  \leftarrow  t_1, \dots, y_{j-1}  \leftarrow  t_{j-1}]
    \equiv \\ 
        x'_1 \colon A'_1, \dots, x'_n \colon A'_n
        \vdash 
        s_j \colon B'_j[y'_1  \leftarrow  s_1, \dots, y_{j-1}  \leftarrow  s_{j-1}].
\end{multline*}

It follows that $C(U)$ is itself a contextual category.

\begin{definition}
    If $\mathcal{C}$ and $\mathcal{C}'$ are contextual categories, then a \textbf{contextual functor}
    is a functor \cite{maclane1971} $F \colon \mathcal{C} \to \mathcal{C}'$ such that:
    \begin{enumerate}
        \item $F(1) = 1$ and if $A \lhd B$ in $\mathcal{C}$, then $F(A) \lhd F(B)$ in $\mathcal{C}'$;
        \item for every object $A$ of $\mathcal{C}$, $F(p(A)) = p(F(A))$;
        \item for all $f$ and $B$ such that $f^*B$ is defined in $\mathcal{C}$,
            \[
                F(f^*B) = F(f)^*F(B) 
                    \quad \text{ and } \quad 
                F(q(f, B)) = q(F(f), F(B)).
            \]
    \end{enumerate}
\end{definition}
$\mathbf{Con}$ denotes the category of contextual categories and contextual functors.

\section{Properties of Contextual Categories}

In this section we state some properties about contextual categories that will be used later.

% generalized projection morphism
The first useful result is about projections; Although we defined projections on pairs $A \lhd B$, 
we might extend the notion to pairs $A \leq B$ defined as \[
    p(B, A) = p(X_1) \circ \dots \circ p(X_n) \circ p(B),
\] where $X_1, \dots, X_n$ is the unique sequence of objects in $\mathcal{C}$ such that
$A \lhd X_1 \lhd \dots \lhd X_n \lhd B$ in $\mathcal{C}$. 
In the case where $A = B$, we take $p(B, A) = \id_A$.
Notation: $B \openrightarrow A$. % TODO: fix notation?

In type theory, proving a theorem often requires working under a set of assumptions (a context).
In categorical semantics, "assuming a context $A$" corresponds to taking the relativization of $\mathcal{C}$
to $A$, noted as $\mathcal{C}_A$.
More formally, let $\mathcal{C}$ be a contextual category and $A$ an arbitrary object of $\mathcal{C}$. Then we define 
the \textit{relativization} of $\mathcal{C}$ to $A$ as the contextual category
where:
\begin{enumerate}
    \item The objects of $\mathcal{C}_A$ are the objects $B$ of $\mathcal{C}$ such that $A \leq B$.
    \item For objects $B, C$ in $\mathcal{C}_A$, a morphism from $B$ to $C$ in $\mathcal{C}_A$, also referenced as a morphism from $B$ to $C$ over $A$,
          is a morphism $f \colon B \to C$ in $\mathcal{C}$ such that $p(C, A) \circ f = p(B, A)$. Composition of morphisms is inherited from $\mathcal{C}$.
    \item $\parent_A(B) = \parent(B)$ if $A < B$, and $\parent_A(A) = A$.
    \item For objects $B > A$, $p_A(B) = p(B) \colon B \to \parent(B)$.
    \item For $B, C, D$ objects of $\mathcal{C}_A$ with $\parent_A(D) = C$ and morphisms $f \colon B \to C$, we write
          \[
                f^{*_A}D = f^*D \quad \text{ and } \quad q_A(f, D) = q(f, D).
          \]
\end{enumerate}

Now we see that any morphism $f \colon B \to A$ in $\mathcal{C}$ induces a pullback functor 
$f^* \colon \mathcal{C}_A \to \mathcal{C_B}$ which is a contextual functor. The proof of the following 
theorems is included in \cite{streicher}.
This functor can be thought as a way of embedding the knowledge over the context $A$ onto the context $B$.

\begin{lemma}
    Let $\mathcal{C}$ be a contextual category, and $f \colon B \to A$ a morphism in $\mathcal{C}$.
    Then, for any object $D$ with $D \geq A$, we define an object $f^*D \geq B$ and a morphism 
    $q(f, D) \colon f^*D \to D$ by induction over $\level D - \level A$:
    \begin{enumerate}
        \item if $D = A$, \[
                f^*D = B \quad \text{ and } \quad q(f, D) = f;
              \]
        \item if $D > A$, \[
                f^*D = q(f, \parent(D))^*D
                    \quad \text{ and } \quad 
                q(f, D) = q(q(f, \parent(D)), D).
              \]
    \end{enumerate}
    Then, if $D \geq A$, the pair $(p(f^*D, B), q(f, D))$ is a pullback cone over $(f, p(D, A))$.
\end{lemma}
\begin{theorem}
    Let $\mathcal{C}$ be a contextual category and $f \colon B \to A$ a morphism in $\mathcal{C}$.
    By mapping an object $C$ of $\mathcal{C}_A$ to $f^*C$ in $\mathcal{C}_B$,
    and mapping $g \colon C \to D$ over $A$ to $f^*g \colon f^*C \to f^*D$ over $B$
    determined by
    \begin{align*}
        p(f^*D, B) \circ f^*g &= p(f^*C, B) \\
           q(f, D) \circ f^*g &= q(f, C)
    \end{align*}
    we obtain a contextual functor $f^* \colon \mathcal{C}_A \to \mathcal{C}_B$
    called \textit{reindexing along} $f$.
\end{theorem}

\begin{theorem}
    Let $\mathcal{C}$ be a contextual category and $f \colon B \to A$, $g \colon C \to B$ be morphisms in $\mathcal{C}$.
    Then, \[
        g^* \circ f^* = (f \circ g)^* 
            \quad \text{ and } \quad
        q(f \circ g, D) = q(f, D) \circ q(g, f^*D)
    \]
    for objects $D \geq A$.
\end{theorem}

\begin{definition} \cite{streicher}
    Let $\mathcal{C}$ be a contextual category, $A$ and $B$ be objects of $\mathcal{C}$ such that $A \lhd B$.
    The set of \textbf{sections} of $B$ is defined as 
    \begin{equation*}
        \Sections(B) \coloneqq \{ s \colon A \to B   \leftarrow  p(B) \circ s = \id_A\}.
    \end{equation*}
\end{definition}
If $\level(B) = 1$, then 
\[
    \Sections(B) = \{ s  \leftarrow  s \colon 1 \to B \},
\]
as any morphism $s \colon 1 \to B$ satisfies $p(B) \circ s = \id_1$ since
$1$ is a terminal object. In this case, the notion of section
matches the notion of \textit{global element} of $B$. 
The concept of section of a context $\langle x_1 \colon A_1, \dots, x_n \colon A_n, y \colon B \rangle$ is 
the categorical equivalent of a term of type $B$ under context $\langle x_1 \colon A_1, \dots, x_n \colon A_n \rangle$.

\section{Relevance of the Context Category} \label{sect:relevance-of-the-context-category}

The context category is the first example of what it is called a \textbf{term model}.
This means that it is a category built directly from the syntax of a theory 
without picking any particular mathematical interpretation. 
This gives us the \textit{most generic} model of that theory.
The three main properties of this models are the following:
\begin{enumerate}
    \item \textbf{Soundness}: Any model of such theory corresponds to a structure preserving 
          functor out of the term model.
    \item \textbf{Completeness}: A judgement is provable/derivable in the theory if it holds in all models, 
          or equivalently if it holds in the term model.
    \item \textbf{Initiality}: The term model is often the initial object of the category of all models.
\end{enumerate}
Because of completeness, navigating $C(U)$ is equivalent to manipulating the syntax of $U$.

We start by proving the completeness of $C(U)$. The following completeness theorem is 
direct from the definition of $C(U)$:
\begin{theorem}[Completeness Theorem] \label{thm:completeness-context-category}
    Let $\mathcal{M}$ be the canonical interpretation of a theory $U$ in $C(U)$.
    \begin{enumerate}
        \item The $T$-rule \[
                \dfrac{
                    x_1 \colon A_1, \dots, x_n \colon A_n
                }{
                    A \type
                }
            \] is derivable if and only if
            $x_1 \colon A_1, \dots, x_n \colon A_n, x \colon A$
            is a context, i.e., if $[x_1 \colon A_1, \dots, x_n \colon A_n, x \colon A]$
            in $C(U)$.
        \item The $\in$-rule \[
                \dfrac{
                    x_1 \colon A_1, \dots, x_n \colon A_n
                }{
                    t \colon A
                }
            \]  is derivable if and only if
            $\langle x_1, \dots, x_n, t \rangle$ is a section of 
            $[x_1 \colon A_1, \dots, x_n \colon A_n, x \colon A]$
            in $C(U)$.
        \item If the following $T$-rules are derivable: \[
                \Gamma \vdash A_1 \type 
                    \quad \text{ and } \quad
                \Gamma \vdash A_2 \type,
            \] then the rule \(
                \Gamma \vdash A_1 = A_2
            \) is derivable if and only if \[
                [\Gamma, x \colon A_1]
                = 
                [\Gamma, x \colon A_2],
            \]
            where $x$ is a variable that does not appear in $\Gamma$.
        \item If the following $\in$-rules are derivable: \[
                \Gamma \vdash t \colon A 
                    \quad \text{ and } \quad
                \Gamma \vdash s \colon A,
            \] then the rule $\Gamma \vdash t = s \colon A$ is derivable
            if, and only if, $\langle x_1, \dots, x_n, t \rangle$ and 
            $\langle x_1, \dots, x_n, s \rangle$
            are equal sections of $[\Gamma, x: A]$ in $C(U)$,
            where $x$ is a variable not appearing in $\Gamma$.
    \end{enumerate}
\end{theorem}

Now we prove the initiality of $C(U)$, a special case of Voevodsky's initiality conjecture \cite{voevodsky2017}. 
Hence, we need to define interpretation of a theory $U$
in a contextual category, and see that the interpretation into $C(U)$ is the 
initial object of the category of interpretations of $U$.

Given a contextual category $\mathcal{C}$ and a theory $U$, an \textit{interpretation} of $U$ in $\mathcal{C}$
is a mapping $\mathcal{M}$ from the $T$- and $\in$-rules of $U$ to the $\mathcal{C}$ such that
\begin{enumerate}
    \item if $R$ is a derived $T$-rule \(
            \dfrac{
                x_1 \colon A_1, \dots, x_n \colon A_n
            }{
                A \text{ is a type}   
            }
        \) and we define rules \\ 
        \(
            R_i \equiv \dfrac{
                x_1 \colon A_1, \dots, x_{i-1} \colon A_{i-1}
            }{
                A_i \text{ is a type}   
            }
        \) for $1 \leq i \leq n$, then $\mathcal{M}[R_1], \dots, \mathcal{M}[R_n], \mathcal{M}[R]$ are
        objects of $\mathcal{C}$ such that \[
            1 \lhd \mathcal{M}[R_1] \lhd \dots \lhd \mathcal{M}[R_n] \lhd \mathcal{M}[R]
        \]
        in $\mathcal{C}$.

    \item if $R_t$ is a derived $\in$-rule \(
            \dfrac{
                x_1 \colon A_1, \dots, x_n \colon A_n   
            }{
                t \colon A
            }
        \) then $\mathcal{M}[R_t]$ is section of $\mathcal{M}[R]$.

    \item if $\Gamma \vdash A = A'$ is a derived $T=$rule, then if \[
            R \equiv \Gamma \vdash A \type
            \quad \text{ and } \quad
            R' \equiv \Gamma \vdash A' \type,
        \] then $\mathcal{M}[R] = \mathcal{M}[R']$.
    \item if $\Gamma \vdash t = t' \colon A$ is a derived $\in=$rule, then if \[
            R_t \equiv \Gamma \vdash t \colon A, \quad
            R_{t'} \equiv \Gamma \vdash t' \colon A, \quad
            R \equiv \Gamma \vdash A \type,
        \] then both $\mathcal{M}[R_t]$ and $\mathcal{M}[R_{t'}]$ are sections of $\mathcal{R}$.
        and $\mathcal{M}[R_t] = \mathcal{M}[R_{t'}]$
\end{enumerate}

There is a canonical interpretation $\mathcal{M}$ of $U$ in $\mathcal{C}(U)$, defined as follows:
\begin{enumerate}
    \item if $R \equiv x_1 \colon A_1, \dots, x_n \colon A_n \vdash A \type$ is a derived $T$-judgement, 
            then we define \[
                \mathcal{M}[R] = [x_1 \colon A_1, \dots, x_n \colon A_n, x_{n+1} \colon A];
            \]
    \item if $R_t \equiv x_1 \colon A_1, \dots, x_n \colon A_n \vdash t \colon A$ is a derived $\in$-judgement,
            then we define \[
                \mathcal{M}[R_t] = [\langle x_1, \dots, x_n, t \rangle]
            \] a section of $\mathcal{M}[R]$.
\end{enumerate}
The Completeness Theorem \ref{thm:completeness-context-category} guarantees that $\mathcal{M}$ is well defined.

Now we shall define the concept of homomorphism between interpretations of a theory $U$.
Let $\mathcal{C}$, $\mathcal{C}'$ be two contextual categories, and let $\mathcal{M_C}$ and $\mathcal{M_{C'}}$
be two $U$-interpretations in $\mathcal{C}$ and $\mathcal{C}'$ respectively. Then, an interpretation 
homomorphism is a contextual functor $F \colon \mathcal{C} \to \mathcal{C}'$ such that:
\begin{enumerate}
    \item if $R$ is a $T$-rule, then $F \mathcal{M_C}[R] = \mathcal{M_{C'}}[R]$;
    \item if $R_t$ is a $\in$-rule, then $F \mathcal{M_C}[R_t] = \mathcal{M_{C'}}[R_t]$.
\end{enumerate}

So we can define the category $\mathbf{Con}_U$ as the category with pairs of contextual categories and $U$-interpretations
and their homomorphisms. We can easily see that $(C(U), \mathcal{M})$ is an
initial object of $\mathbf{Con}_U$. Consider an object $(\mathcal{C}, \mathcal{N})$ of
$\mathbf{Con}_U$. Let construct a morphism $F \colon ((C(U), \mathcal{M})) \to (\mathcal{C}, \mathcal{N})$, i.e., we have to construct
a contextual functor $F \colon C(U) \to \mathcal{C}$ that \textit{commutes} with the interpretation functions.
If $[\Gamma]$ an object of $C(U)$, then there is a $T$-rule $R^\Gamma$ such that $\mathcal{M}[R^\Gamma] = \Gamma$. We define 
$F \Gamma = \mathcal{N}[R^\Gamma]$. 
In order to define $F$ on the morphisms of $C(U)$, it suffices to define it for the sections of $C(U)$, as any morphism 
\[
    \langle t_1, \dots, t_m \rangle \colon
    \langle x_1 \colon A_1, \dots, x_n \colon A_n \rangle \to
    \langle y_1 \colon B_1, \dots, y_m \colon B_m \rangle
\]
can be built by composition and pullback functors of sections and projections, and the latter is uniquely determined by the condition that $F$
is a contextual functor.
Hence, given a section $s = \langle x_1, \dots, x_n, t \rangle$ of $R \equiv x_1 \colon A_1, \dots, x_n \colon A_n, x \colon A$,
by definition there exists a derived $\in$-rule $R_t$ such that $\mathcal{M}[R_t] = s$, so we can define
$F s = \mathcal{N}[R_t]$.
The uniqueness of $F$ is rather simple to see, as we cannot assign any other values to any object or morphism of $C(P)$ without 
conflicting with the structural conditions of the functor.
Hence, $(C(U), \mathcal{M})$ is an initial object of $\mathbf{Con}_U$.

Soundness comes from the initiality of $C(U)$, as any model of $U$, i.e., any element of $\mathbf{Con}_U$, 
is equivalent to the initial morphism $F$ we just defined.

\section{Definition of the Problem}

In this section we introduce a formal definition of our 
problem, proposed at the beginning of the thesis,
and reformulate it to an equivalent problem easier to solve
with artificial intelligence.

Our problem might be formalized in the following manner:
Given a theory $U$ and a derived rule of $U$ of the form 
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        A \type
    }
\end{equation*}
we want to find an expression $t$ of $U$ such that the rule 
\begin{equation*}
    \dfrac{
        x_1 : A_1, \dots, x_n : A_n
    }{
        t : A 
    }
\end{equation*}
is a derived rule of $U$.

This problem is undecidable, in general, as it follows from the undecidability
of the inhabitation in a $GAT$. This is why we will not try to find
an algorithm that either finds a term or says that the type is empty.
We aim to find an algorithm that, in case that the type is 
inhabited, it finds a term in a finite number of steps.
In case that the type is empty, it does not have to give an
output.

We can consider this problem from the point of view of the 
context category $C(U)$, where by the Completeness Theorem \ref{thm:completeness-context-category}
it is rephrased as follows:
Given a context $\langle x_1 : A_1, \dots, x_n : A_n \rangle$
in $C(U)$, we want to find a section of $\langle x_1 : A_1, \dots, x_n : A_n, x : A \rangle$
in $C(U)$.

By naming $\Gamma \equiv \langle x_1 : A_1, \dots, x_n : A_n \rangle$,
we can restrict our search to the \textit{relativization} of $C(U)$ to $[\Gamma]$,
that we denote as $\mathcal{C}_\Gamma$.
We denote a morphism $\langle x_1, \dots, x_n, t_{n+1}, \dots, t_{n+m} \rangle$  in $\mathcal{C}_\Gamma$,
as $\langle t_{n+1}, \dots, t_{n+m} \rangle_\Gamma$.

\section{Expansion Algorithm}

In this section, we define the \textit{expansion algorithm}
that will be responsible of finding $\langle t \rangle_\Gamma$
in $\mathcal{C}_\Gamma$.
The complexity of this algorithm is really high, and it will serve
as a baseline to be optimized by a graph neural network. We will
explain this in the last chapter.

First, consider a finite subcategory $G$ of $\mathcal{C}_A$ which 
contains $\Gamma$, $\Gamma, x \colon A$ and
its projection.
The process consists of the construction of a sequence $\{ G_i\}_i$
of finite subcategories of $\mathcal{C}_\Gamma$ such that $G_0 = G$,
and that $G_{n+1}$ is computed by the process described below
from $G_n$, and where $G_n \subseteq G_{n+1}$ for all $n \geq 0$.
We say that an object or morphism of $\mathcal{C_A}$ is \textit{constructible by expansion}
if there exists an $n \geq 0$ such that the given object or morphism 
is contained in $G_n$. 
The process of constructing $G_{n+1}$ from $G_n$ consists of applying
the following rules to each context 
$\Gamma y_1 \colon B_1, \dots, y_m \colon B_m$
of $G_n$:
\begin{itemize}
    \ruleitem{E1}{sort-symbol-intro} For each sort symbol $S$ with a well-formed
        introductory rule \[
            \dfrac{
                z_1 \colon C_1, \dots, z_n \colon C_w
            }{
                A(z_1, \dots, z_w) \type
            }
        \] such that $G_n$ contains sections
        \begin{align*} % TODO: fix alignment
            \Gamma, y_1 : B_1, \dots, y_m : B_m
            &\xrightarrow{\langle y_1, \dots, y_m, t_1 \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z_1 : C_1 \\
            \Gamma, y_1 : B_1, \dots, y_m : B_m
            &\xrightarrow{\langle y_1, \dots, y_m, t_2 \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z_2 : C_2[z_1  \leftarrow  t_1] \\
            \vdots \\
            \Gamma, y_1 : B_1, \dots, y_m : B_m
            &\xrightarrow{\langle y_1, \dots, y_m, t_w \rangle_\Gamma}
            \Gamma y_1 : B_1, \dots, y_m : B_m, z_w : C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}] \\
        \end{align*}
        then we add the context
        $\Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, y \colon A(t_1, \dots, t_w)$ 
        and its projection to $G_{n+1}$.
    \ruleitem{E2}{variable-duplication} Add the morphisms \[
            \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m 
            \xrightarrow{\langle y_1, \dots, y_m, x_i\rangle_\Gamma}
            \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, x \colon A_i 
        \] \[
            \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m
            \xrightarrow{\langle y_1, \dots, y_m, y_j \rangle_\Gamma}
            \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, y \colon B_j
        \]
        
        and their corresponding projections for each $1 \leq i \leq n$ and each $1 \leq j \leq m$.
    \ruleitem{E3}{operator-symbol-intro} For every operator symbol $F$ with a well-formed
        introductory rule \[
            \dfrac{
                z_1 \colon C_1, \dots, z_w \colon C_w
            }{
                F(z_1, \dots, z_w) \colon C
            }
        \] if the following sections exist in $G_n$
        \begin{align*} % TODO: fix alignment
            \Gamma, y_1 : B_1, \dots, y_m : B_m 
            &\xrightarrow{\langle y_1, \dots, y_m, t_1 \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z_1 : C_1  \\
            \Gamma, y_1 : B_1, \dots, y_m : B_m 
            &\xrightarrow{\langle y_1, \dots, y_m, t_2 \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z_2 : C_2[z_1  \leftarrow  t_1]  \\
            \vdots \\
            \Gamma, y_1 : B_1, \dots, y_m : B_m 
            &\xrightarrow{\langle y_1, \dots, y_m, t_w \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z_w : C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}] \\
        \end{align*}
        then we add the section
        \[ % TODO: fix alignment
            \Gamma, y_1 : B_1, \dots, y_m : B_m 
            \xrightarrow{\langle y_1, \dots, y_m, F(t_1, \dots, t_w) \rangle_\Gamma}
            \Gamma, y_1 : B_1, \dots, y_m : B_m, z : C[z_1  \leftarrow  t_1, \dots, z_w  \leftarrow  t_w]
        \]
    \ruleitem{E4}{pullback} For every morphism
        \[
            \langle t_1, \dots, t_{m-1} \rangle_\Gamma
            : \Gamma, z_1 : C_1, \dots, z_w : C_w 
            \to \Gamma, y_1 : B_1, \dots, y_{m-1} : B_{m-1}
        \]
        we construct the corresponding pullback % TODO: fix alignment
        \[
            \begin{tikzcd}
                \Gamma, z_1: C_1, \dots, z_w : C_w, y_m : B_m[t_1  \leftarrow  y_1, \dots, t_{m-1}  \leftarrow  y_{m-1}]
                \arrow[r, "\langle t_1{,} \dots{,} t_{m-1}{,} y_m \rangle_\Gamma"] \arrow[d, open] 
                & \Gamma, y_1 : B_1, \dots, y_m : B_m
                \arrow[d, open] \\
                \Gamma, z_1 : C_1, \dots, z_w : C_w
                \arrow[r, "\langle t_1 {,} \dots {,} t_{m-1} \rangle_\Gamma"]
                & \Gamma, y_1 : B_1, \dots, y_{m-1} : B_{m-1}
            \end{tikzcd}
        \]
        in $G_{n+1}$.
\end{itemize} 

Now we will prove that every object and every morphism of $\mathcal{C}_A$ are constructible by expansion.
But to do so, we will introduce the concept of $\depth$ of an expression in order to 
prove a the result by induction. This definition is inspired to the one in the next chapter taken from \cite{streicher}.

\begin{definition}
    By structural induction of expressions we define a function \textit{depth} mapping 
    the set of expressions and premisses to $\omega$:
    \begin{enumerate}
        \item if $x$ is a variable, then $\depth[x] = 1$;
        \item if $A$ is a sort symbol, and $t_1, \dots, t_n$ are expressions, then \[
            \depth[A(t_1, \dots, t_n)] = 1 + \sum_{i = 1}^n \depth[t_i];
        \]
        \item if $F$ is an operator symbol, and $t_1, \dots, t_n$ are expressions, then \[
            \depth[F(t_1, \dots, t_n)] = 1 + \sum_{i=1}^n \depth[t_i];
        \]
        \item if $\Gamma \equiv x_1 \colon A_1, \dots, x_n \colon A_n$ is a premise, then \[
            \depth[\Gamma] = \sum_{i = 1}^n \depth[A_i].
        \]
    \end{enumerate}
\end{definition}

\begin{lemma}
    Every object and every section of $\mathcal{C}_\Gamma$ are constructible by expansion.
\end{lemma}
\begin{proof}
    We perform this proof by induction over the $\degree$ of the objects and sections, where
    \begin{enumerate}
        \item if $\Delta$ is a context over $\Gamma$, then \[
            \degree[\Delta] = \depth[\Delta] - \depth[\Gamma];
        \]
        \item if $\Delta \equiv \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m$ is a context and 
            $[\langle y_1, \dots, y_m, t \rangle_\Gamma]$ is a section of $[\Delta, x \colon A]$, then \[
                \degree[\langle y_1, \dots, y_m, t \rangle_\Gamma] = \degree[\Delta] + \depth[t].
            \]
    \end{enumerate}
    Together with a secondary induction on derivation length.

    Consider then a context $\Delta \equiv \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m$. We want to see that $[\Delta]$ is constructible by expansion.
    As $\Delta$ is a context, if we denote $\Delta' \equiv \Gamma, y_1 \colon B_1, \dots, y_{m-1} \colon B_{m-1}$, we have that \[
        \Delta' \vdash B_m \type
    \] is a derived rule of $U$. This implies that $\Delta'$ is itself a context, and by induction hypothesis, as \[
        \degree[\Delta] = \degree[\Delta'] + \depth[B_m] \geq \degree[\Delta'] + 1 > \degree[\Delta'],
    \] we have that $[\Delta']$ is constructible by expansion.
    Moreover, as the rule $\Delta' \vdash B_m \type$ can only be derived by \ref{rule:cf2a}, there exists a sort symbol $S$
    with a well-formed introductory rule \[
        z_1 \colon C_1, \dots, z_w \colon C_w \vdash A(z_1, \dots, z_w) \type
    \] and derived rules
    \begin{align*}
        \Delta' \vdash t_1 \colon& C_1 \\
        \Delta' \vdash t_2 \colon& C_2[z_1  \leftarrow  t_1] \\
        & \vdots \\
        \Delta' \vdash t_w \colon& C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}]
    \end{align*}
    such that $B_m = A(t_1, \dots, t_m)$ as expressions. 
    This means that these are sections:
    \begin{align*}
        [\langle y_1, \dots, y_m, t_1 \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_1 \colon C_1] \\
        [\langle y_1, \dots, y_m, t_2 \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_2 \colon C_2[z_1  \leftarrow  t_1]] \\
        &\vdots \\
        [\langle y_1, \dots, y_m, t_w \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_w \colon C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}]] 
    \end{align*}
    And clearly, as \begin{multline*}
        \degree[\Delta] = \degree[\Delta'] + \depth[A(t_1, \dots, t_m)] > \\ 
        > \degree[\Delta'] + \depth[t_j] = \degree[\langle y_1, \dots, y_m \rangle_\Gamma]
    \end{multline*}
    for every $j$, with $1 \leq j \leq w$, they are all constructible by expansion by induction hypothesis.
    Finally, by applying the expansion rule \ref{rule:sort-symbol-intro}, we have that $[\Delta]$ is constructible by expansion.
    
    Now, consider a section $[\langle y_1, \dots, y_m, t \rangle_\Gamma]$ of $[\Delta, x \colon A]$, hence the rule 
    \begin{equation}
        \Delta \vdash t \colon A \label{eq:underlying-rule-operator}
    \end{equation} 
    is a derived rule of $U$. We have that $\degree[\langle y_1, \dots, y_m, t \rangle_\Gamma] = \degree[\Delta] + \depth[t]$, hence 
    by induction hypothesis $[\Delta]$ is constructible by expansion. 
    We now study how the rule \eqref{eq:underlying-rule-operator} was derived. 
    As it is a $\in$-rule, it must have been derived by either \ref{rule:t1}, \ref{rule:cf1} or \ref{rule:cf2a}.
    We treat each case separately:
    \begin{enumerate}
        \item[\ref{rule:t1}] Then, there exists an expression $A'$ such that both $\Delta \vdash t \colon A'$ and $\Delta \vdash A = A'$ are derived rules of $U$.
            The latter rule gives us that $[\Delta, x \colon A] = [\Delta, x \colon A']$, hence we have to see that the section $[\langle y_1, \dots, y_m, t \rangle_\Gamma]$ of $[\Delta, x \colon A']$ is constructible by expansion.
            As the length of the derivation of $\Delta \vdash t \colon A'$ is at least one \ref{rule:t1} application shorter that the derivation length of $\Delta \vdash t \colon A$, by secondary induction, we have that the section is constructible by expansion.
        \item[\ref{rule:cf1}] Then, there exists an $i$, with $1 \leq i \leq n$, such that $t = x_i$ and $A = A_i$;
            or there exists a $j$, with $1 \leq j \leq m$, such that $t = y_j$ and $A = B_j$.
            We just prove the first case as the other proof is equivalent. So as $[\Delta]$ is constructible by expansion,
            by applying the expansion rule \ref{rule:variable-duplication} to it, we get that the section \[
                [\langle y_1, \dots, y_m, x_i \rangle_\Gamma] \colon [\Delta] \to [\Delta, x \colon A]
            \] is constructible by expansion.
        \item[\ref{rule:cf2a}]  Then, there is an operator symbol $F$, with a well-formed introductory rule \[
                z_1 \colon C_1, \dots, z_w \colon C_w \vdash F(z_1, \dots, z_w) \colon B
            \] and derived rules:
            \begin{align*}
                \Delta \vdash t_1 \colon& C_1 \\
                \Delta \vdash t_2 \colon& C_2[z_1  \leftarrow  t_1] \\
                &\vdots \\
                \Delta \vdash t_w \colon& C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}] 
            \end{align*}
            such that $t = F(t_1, \dots, t_w)$ and $A = B[z_1  \leftarrow  t_1, \dots, z_w  \leftarrow  t_w]$.
            Then, as 
            \begin{multline*}
                \degree[\langle y_1, \dots, y_m, F(t_1, \dots, t_m) \rangle_\Gamma] = \degree[\Delta] + \depth[F(t_1, \dots, t_w)] > \\
                    > \degree[\Delta] + \depth[t_j] = \degree[\langle y_1, \dots, y_m, t_j \rangle_\Gamma]
            \end{multline*}
            for all $j$, with $1 \leq j \leq w$, by the induction hypothesis, the following sections are constructible by expansion: 
            \begin{align*}
                [\langle y_1, \dots, y_m, t_1 \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_1 \colon C_1] \\
                [\langle y_1, \dots, y_m, t_2 \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_2 \colon C_2[z_1  \leftarrow  t_1]] \\
                &\vdots \\
                [\langle y_1, \dots, y_m, t_w \rangle_\Gamma] &\colon [\Delta'] \to [\Delta', z_w \colon C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}]] 
            \end{align*}
            Then by applying the expansion rule \ref{rule:operator-symbol-intro}, we get that the section \[
                [\langle y_1, \dots, y_m, t \rangle_\Gamma] \colon [\Delta] \to [\Delta, x \colon A]
            \] is constructible by expansion.
    \end{enumerate}
\end{proof}

\begin{theorem} \label{thm:exhaustivity-for-gats}
    Every object and every morphism of $\mathcal{C}_A$ are constructible by expansion.
\end{theorem}
\begin{proof} 
    From the last lemma we have that both objects and sections are constructible by expansion.
    We now have to prove that any morphism is constructible by expansion.
    Let $[\langle t_1, \dots, t_m \rangle_\Gamma]: [\Gamma, y_1 \colon B_1, \dots, y_m \colon B_m] \to [\Gamma, z_1 \colon C_1, \dots, z_w \colon C_w]$
    be a morphism on $\mathcal{C}_\Gamma$.  We will denote $\Delta \equiv \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m$.
    Then we know that the following rules
    \begin{align*}
        \Delta \vdash t_1 \colon& C_1 \\
        \Delta \vdash t_2 \colon& C_2 [z_1  \leftarrow  t_1] \\ 
        & \vdots \\ 
        \Delta \vdash t_w \colon& C_2 [z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}]
    \end{align*} 
    are derivable rules of $U$. So, by applying the lemma, we have that the following sections are constructible by expansion: 
    \begin{align*}
        [\langle y_1, \dots, y_m, t_1 \rangle_\Gamma] \colon [\Delta] &\to [\Delta, z_1 \colon C_1] \\
        [\langle y_1, \dots, y_m, t_2 \rangle_\Gamma] \colon [\Delta] &\to [\Delta, z_2 \colon C_2[z_1  \leftarrow  z_2]] \\
        & \vdots \\
        [\langle y_1, \dots, y_m, t_w \rangle_\Gamma] \colon [\Delta] &\to [\Delta, z_w \colon C_w[z_1  \leftarrow  t_1, \dots, z_{w-1}  \leftarrow  t_{w-1}]]
    \end{align*} 

    Now we show that the morphism 
    \[
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m \rangle
        \xrightarrow{\langle y_1, \dots, y_m, t_1, \dots, t_w \rangle_\Gamma}
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, z_1 \colon C_1, \dots, z_w \colon C_w
    \]
    is constructible by expansion. To do this, we  prove that, given $j$ with $2 < j \leq n$, if
    \[
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m 
        \xrightarrow{\langle y_1, \dots, y_m, \dots, t_{j-1} \rangle_\Gamma}
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, z_1 \colon C_1, \dots, z_{j-1} \colon C_{j-1}
    \]
    \[
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m
        \xrightarrow{\langle y_1, \dots, y_m, t_j \rangle_\Gamma}
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, z_j \colon C_j[z_1  \leftarrow  \dots z_{j-1}  \leftarrow  t_{j-1}]
    \]
    are constructible by expansion so is
    \[
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m
        \xrightarrow{\langle y_1, \dots, y_m, t_1, \dots, t_j \rangle_\Gamma}
        \Gamma, y_1 \colon B_1, \dots, y_m \colon B_m, z_1 \colon C_1, \dots, z_j \colon C_j 
    \]
    By the previous lemma, the object $[\Delta, z_1 \colon C_1, \dots, z_j \colon C_j]$ is constructible by expansion,
    and by applying the expansion rule \ref{rule:pullback} on it, the following diagram is constructible by expansion:
    \[
    \begin{tikzcd} % TODO: fix overflow
        \Delta, z_j : C_j[t_1  \leftarrow  z_1, \dots, t_j  \leftarrow  z_j]
            \arrow[r, "\langle y_1{,} \dots{,} y_m{,} t_1{,} \dots {,} t_{j-1} {,} z_j \rangle_\Gamma"]
            \arrow[d, open] 
        & \Delta, z_1 : C_1, \dots, z_j : C_j 
            \arrow[d, open] 
        \\
        \Delta
            \arrow[r, "\langle y_1{,} \dots{,} y_m{,} t_1{,} \dots{,} t_{j-1} \rangle_\Gamma"] &
        \Delta, z_1 : C_1, \dots, z_{j-1} : C_{j-1}
    \end{tikzcd}
    \]
    and by composition we get that 
    \[
        \langle y_1, \dots, y_m, t_1, \dots, t_{j-1}, z_j \rangle_\Gamma
        \circ
        \langle y_1, \dots, y_m, t_j \rangle_\Gamma
        = 
        \langle y_1, \dots, y_m, t_1, \dots, t_j \rangle_\Gamma
    \]
    is constructible by expansion.
    It remains to see that the morphism
    \[
        \Gamma,
            y_1 : B_1, \dots, y_m : B_m,
            z_1 : C_1, \dots, z_w : C_w
        \xrightarrow{\langle z_1, \dots, z_w \rangle_\Gamma}
        \Gamma,
            z_1 : C_1, \dots, z_w : C_w
    \]
    is constructible by expansion. That comes directly from the 
    fact that $[\Gamma, z_1 \colon C_1, \dots, z_w \colon C_w]$
    is constructible by expansion, and repeated application of the rule \ref{rule:pullback}
    gives us the following diagram: 
    \[
    \begin{tikzcd}
        \Gamma, 
        y_1 : B_1, \dots, y_m : B_m
        z_1 : C_1, \dots, z_w : C_w
            \arrow[r, "\langle z_1{,} \dots{,} z_w \rangle_\Gamma"]
            \arrow[d, open] 
        & \Gamma,
        z_1 : C_1, \dots, z_w : C_w
            \arrow[d, open] 
        \\
        \Gamma,
        y_1 : B_1, \dots, y_m : B_m
        z_1 : C_1, \dots, z_{w-1} : C_{w-1}
            \arrow[r, "\langle z_1{,} \dots{,} z_{w-1} \rangle_\Gamma"]
            \arrow[d, open] 
        &
        \Gamma,
        z_1 : C_1, \dots, z_{w-1} : C_{w-1}
            \arrow[d, open] 
        \\
        \vdots 
            \arrow[d, open] 
        & \vdots 
            \arrow[d, open]
        \\
        \Gamma, 
        y_1 : B_1, \dots, y_m : B_m,
        z_1 : C_1
            \arrow[r, "\langle z_1 \rangle_\Gamma"]
            \arrow[d, open] 
        & \Gamma, z_1 : C_1
            \arrow[d, open] 
        \\
        \Gamma,
        y_1 : B_1, \dots, y_m : B_m
            \arrow[r, "\langle \rangle_\Gamma"] 
        & \Gamma
    \end{tikzcd}
    \]
    This completes the proof of Theorem \ref{thm:exhaustivity-for-gats}.
\end{proof}